Quantitative Benchmarking for Automatically Generated Conjectures

### 2018-03-15 TEBenchmarkPaper/article.tex
Fix up table spacing
Fix table border
Lots of additions and changes

% Look carefully at each system's (paper's) eval section

@@

% Lay out the field of TE: motivation, systems, evaluation, problems...

@@

We attempt to solve this discrepancy, at least for the foreseeable future, by
defining a standard, unambiguous benchmarking approach with which to compare
the conjecture generation of MTE systems. Our contributions include:

\begin{itemize}
\item A general methodology for benchmarking conjecture generation.
\item Resolving the issue of ``interestingness'' through the use of
  theorem-proving benchmarks as a ground-truth.
\item A specific instantiation of this methodology, using the Tons of Inductive
  Problems benchmark as a corpus.
\item Automated tooling to perform this benchmarking.
\item Application of our methodology to the QuickSpec and IsaCoSy MTE systems,
  and a comparison and discussion of the results.
\end{itemize}

@@

A major challenge for both ATF and MTE is choosing how to narrow down
the set of generated conjectures to those deemed ``interesting'', since this is
an imprecise term with many different interpretations. For example, all existing
approaches agree that simple tautologies are ``uninteresting'', but differ when
it comes to more complex statements.  Colton \etal{} give a survey of systems for
theory formation and exploration, and their associated notions of
``interestingness'' for concepts and conjectures~\cite{colton2000notion}. Six
notions are identified, which are applied to conjectures as follows:

@@

{\bf Applicability} depends on the number of models in which a
conjecture holds. The system of \Bagai{} conjectures the non-existence of
objects, and hence favours statements with zero applicability. Other systems
treat applicability as a positive aspect: the more applicable the statement, the
more interesting it is.

{\bf Comprehensibility} depends on the complexity of a statement. Simpler
statements are considered more interesting, hence many of the systems explore
simpler/smaller statements before complex/larger ones, to more efficiently find
those which are interesting.

@@

\begin{table}
  \centering
  \begin{tabular}{ |l|l|c|c|c|c|c|c| }
    \hline
    \multirow{2}{*}{\textbf{Program}}                     &
    \multirow{2}{*}{\textbf{Conjecture Types}}            &
    \multicolumn{6}{c}{\textbf{Interestingness Measures}} \\ \hhline{~~------}
    \tRow{          &                    & \iE & \iN & \iS & \iA & \iC & \iU}
    \tRow{AM        & \tIFF, \tIMP, \tNE &   X &   X &   X &   X &   X &   X}
    \tRow{GT        & \tIFF, \tIMP, \tNE &   X &   X &   X &   X &   X &   X}
    \tRow{Graffiti  & \tINE              &   X &   X &   X &     &   X &   X}
    \tRow{\Bagai{}  & \tNE               &     &   X &     &   X &   X &    }
    \tRow{HR        & \tIFF, \tIMP, \tNE &   X &   X &   X &   X &   X &   X}
    \tRow{QuickSpec & \tEQ               &   X &   X &     &   X &   X &    }
    \tRow{Speculate & \tCON \tEQ / \tINE &   X &   X &     &   X &   X &    }
    \tRow{IsaCoSy   & \tEQ               &   X &   X &     &   X &   X &   X}
    \tRow{IsaScheme & \tEQ               &   X &   X &     &   X &   X &   X}
  \end{tabular}
  \caption{Classification of MTE systems from~\cite{colton2000notion}, extended
    to include those from~\cite{claessen2013automating} (QuickSpec is the
    conjecture generation component of HipSpec) and the similar Speculate
    system~\cite{braquehais2017speculate}. The interestingness measures are
    to include four more recent systems. The interestingness measures are
    \iE{}mpirical plausibility, \iN{}ovelty, \iS{}urprisingness,
    \iA{}pplicability, \iC{}omprehensibility (low complexity) and \iU{}tility.}
  \label{table:colton}
\end{table}

@@

We summarise the criteria in Table 1 and state which were used in key
ATF/MTE systems. The entries for AM, GT, Graffiti, \Bagai{} and HR are based
on this survey (the latter is Colton's own system). We extend this analysis to
some more recent systems in the remaining rows (based on our understanding of
their function): QuickSpec~\cite{QuickSpec},
Speculate~\cite{braquehais2017speculate},
IsaCoSy~\cite{Johansson.Dixon.Bundy:conjecture-generation}
and IsaScheme~\cite{MontanoRivas2011}.

The latter systems are described in more detail in section~\ref{section:mte},
but all use testing to check for counterexamples, which ensures results are
empirically plausible and applicable (conditions are satisfiable, types are
inhabited, etc.). They also use a small-to-large search order, ensuring more
easily-comprehensibile conjectures are explored first.

@@

This diversity of approaches makes it difficult to compare the existing
evaluations of these systems directly. In particular, evaluation methods created
for one system might not make sense for another, due to assumptions made about
the algorithms' operation. For example, the novelty filters used by QuickSpec
and Speculate are applied to the whole output set, guaranteeing that no
special-cases will appear; yet we cannot assume this in the case of IsaCoSy,
since its constraint solver allows special cases if they're found \emph{before}
the general case (it does not go back and discard previous results).

With such ambiguous and varied goals, approaches and assessment criteria it is
difficult to compare MTE systems in a quantitative way, and hence to form some
measure of ``progress'' for the field. Our proposed benchmarking methodology
attempts to remedy this by setting a clear target, hopefully in line with the
myriad goals of existing researchers, whilst being generally applicable to
systems of this type.

@@

In this section we put the Mathematical Theory Exploration approach
into context both historically, and with respect to related approaches
such as in Automated Theory Formation systems.

@@

The first automated mathematical system, the mechanical calculator
(known as the Pascaline), was an adding machine that could perform
additions and subtractions directly and multiplication and divisions
by repetitions, and was conceived by Pascal in 1642 while reorganising
tax revenues~\cite{d'ocagne}. Subsequent early systems include
M\"uller's universal calculating machine in 1784, which he built for
the purpose of calculating and printing numerical tables: he invented
this when he had to check and recalculate some tables relating to the
volumes of trees~\cite[p. 65]{lindgren}. Thirty seven years later,
Babbage invented his famous difference engine: an automatic,
mechanical calculator designed to tabulate polynomial functions. This
was inspired by a flawed table of logarithms and the idea that
machines would be quicker and more reliable~\cite{bowden}. It is
interesting to note background and motivation: while Pascal and
Babbage were mathematicians, with an interest in engineering, M\"uller
was an engineer with an interest in mathematical knowledge. All three
systems were conceived as an aid to mathematicians, as well as
scientists, accountants and surveyors.

@@

We limit our discussion to applications in mathematics, although there are
similar approaches being applied to the task of \emph{scientific} discovery.
Scientific knowledge relies on inductive reasoning and experimental testing,
which (in principle) are not so important to a mathematical theory. However,
these techniques \emph{do} play an important role in many existing formation and
exploration tools: in particular, inductive reasoning allows a system to
conjecture statements which may be too difficult for it to deductively prove;
whilst experimental testing, e.g. by checking universally-quantified statements
against some specific values, is a simple way to quickly discard many obvious
falsehoods generated by powerful but imperfect algorithms (as opposed to, say,
limiting ourselves to verified theorem generators).

\subsection{Automated Theory Formation}

@@

% 4. example systems - AM, HR, ...
When Lenat built the AM system~\cite{lenat:77}, there were systems
which could define new concepts for investigation, such as those
described in~\cite{winston}, and systems which could discover
relationships among known concepts, such as Meta-Dendral~\cite{buchanan:75}. No
system could perform both of these tasks: Lenat saw this as the next step. AM
was designed to both construct new concepts and conjecture relationships between
them; fully automating the cycle of discovery in mathematics. Lenat describes
this as follows:

@@

We believe that ATF and MTE share many goals and would benefit from a
closer alignment: both fields are highly specialised and by joining
forces (while being explicit about differences) the techniques being
developed would have greater impact. This closer alignment is already
beginning to take place. The main commonalities between MTE and ATF
are that both are interested in the same aspects of mathematical
thinking, aiming to automatically construct and evaluate elements of a
mathematical theory, including concepts, conjectures, theorems, axioms
and examples. Both strands contrast themselves with Automated Theorem
Proving. The main historical differences between these two approaches
seem to be that in MTE:
\begin{itemize}
\item the main proponents define themselves as mathematicians (they
  hold a PhD and have experience in research mathematics);
\item the primary motivation is to support mathematicians;
\item systems tend to be user-interactive;
\item systems tend to be specific to mathematics,
\end{itemize}

@@

There are three aspects to consider when we evaluate a conjecture generation
system:

\begin{enumerate}
\item The quality of each conjecture, which is the ``interestingness'' discussed
  above.
\item The quality of the \emph{set} of conjectures produced, used for relative
  interestingness criteria like novelty, as well as to assess consistency.
\item Performance of the system, to find the balance struck between output
  quality and time taken.
\end{enumerate}

We also make the assumption that the system is intended to be \emph{general},
and hence isn't hard-coded or fine-tuned to only perform well in a narrow
domain. Examples where this does \emph{not} hold include the natural number
heuristics of AM (discussed above), as well as machine learning models trained
on some specific dataset.

If we limit ourselves to exploring higher-order theories with inductive types
(a domain useful for analysing software), we do find a direct comparison of
three systems~\cite{claessen2013automating}: HipSpec, IsaCoSy and IsaScheme.
QuickSpec, shown in table \ref{table:colton}, is the conjecture-generating
component of HipSpec.

This comparison uses \emph{precision/recall analysis}: a particular set of
definitions is chosen, such as those in figure \ref{figure:list_theory}, and a
\emph{ground truth} is chosen as the 'ideal' set of conjectures which we would
like an MTE system to produce from these definitions.

Each system is run on these definitions, and their outputs are compared to the
ground truth. To score 100\% on precision and recall, an MTE system must output
all of the conjectures which appear in the ground truth, and nothing else:

\begin{itemize}
\item \emph{Precision} is the proportion of a system's output which appears in
  the ground truth. This penalises systems which output large numbers of
  conjectures in the hope that some turn out to be ``good''.
\item \emph{Recall} is the proportion of the ground truth which appears in the
  system's output. This penalises systems which overly restrict their output to
  avoid generating ``bad'' conjectures.
\end{itemize}

Precision and recall fulfil our second requirement, by directly measuring the
quality of the \emph{set} of conjectures. Their ``interestingness'' criterion is
indirect: asking only whether or not a conjecture appears in the ground truth.
A satisfactory solution to our first requirement hence requires careful choice
of the ground truth set.

This particular analysis takes its ground truths from the standard library of
the Isabelle theorem prover, with one involving the natural numbers, and the
other involving the list functions from figure \ref{figure:list_theory}. Since
the library authors have gone to the effort of stating, proving and including
these theorems in every copy of their software, this is a good indication that
they are useful or important.

One problem with this choice is the small size of these libraries. For example,
the benchmark based on Isabelle/HOL's theory of natural numbers given
in~\cite{Johansson.Dixon.Bundy:conjecture-generation} contains only 4
definitions and a ground truth of 12 theorems. Whilst such benchmarks allow
objective comparisons between different approaches, their narrow scope doesn't
provide much indication of performance in different, especially \emph{novel},
domains.

@@

Our main contribution is a benchmarking methodology, shown in figure
\ref{figure:flow_chart}, which generates both a large definition/ground-truth
corpus (an order of magnitude larger than previous work) and a scalable,
statistical approach to evaluating MTE systems using this corpus. We follow a
precision/recall approach similar to the work described above, with the main
difference being the source of definitions and ground truths: we take existing
problem sets designed for automated \emph{theorem proving}, and adapt their
content for use in the \emph{theory exploration} setting.

@@

Automated theorem proving is an active area of research, with large problem sets
and regular competitions to prove as much as possible, as fast as
possible~\cite{pelletier2002development}. These problem sets are an opportunity
for theory exploration, as their definitions and theorems can be used as a
corpus, in the same way that Isabelle libraries have been used in the past.

@@

\begin{itemize}
\item For each problem, there should be a clear distinction between the theorem
  to be proved and the definitions involved, such that the two can be easily and
  meaningfully separated. This rules out problem sets like those of
  SMT-COMP~\cite{barrett2005smt}, where many problems involve uninterpreted
  functions, whose behaviour is \emph{implicit} in the logical structure of the
  theorem statement but not separately \emph{defined}.
\item Definitions should be \emph{strongly typed}, such that they can be
  translated into the native language of each MTE system (in this work, that
  includes Haskell\iffalse FIXME and Isabelle\fi).
\item The problem set should be relevant to the desired domain. In our case, we
  desire higher-order functions and inductively defined types, which rules out
  first-order languages/logics (such as TPTP~\cite{sutcliffe2009tptp}).
\item Since it will act as our ground truth, the problem set should ideally
  contain every ``interesting'' conjecture involving its included
  definitions. Realistically, we should aim for each definition to appear in
  \emph{many} theorem statements; rather than each problem having unique
  definitions.
\item The problem set should be as large as possible, for robustness of the
  resulting statistics.
\end{itemize}

Once such a problem set has been chosen, we must separate the definitions
referenced by the theorems from the theorem statements themselves. The former
will be used as input to the MTE systems under evaluation, whilst the latter
form a ground truth corpus against which to compare the MTE output.

It is important to ensure that there are no duplicate definitions: we are only
concerned with the \emph{logical} content of the input, not the more arbitrary
aspects like the names of functions. For example, consider a problem set which
includes a commutativity theorem for a \texttt{plus} function, and an
associativity theorem for an \texttt{add} function, where the definitions of
\texttt{plus} and \texttt{add} are $\alpha$-equivalent. We would expect an MTE
system to conjecture commutativity and associativity for \emph{both} functions,
or for \emph{neither} function, since they're logically equivalent. Yet a
na\"ive precision/recall analysis would treat commutativity of \texttt{add} and
associativity of \texttt{plus} as \emph{uninteresting}, since they don't appear
in the ground truth.

@@

We could, in theory, send these de-duplicated definitions straight into an MTE
system, and use the updated theorem statements as the ground truth for analysis.
However, this would cause two problems:

\begin{itemize}
\item The result would be a single data point, which makes it difficult to
  infer performance \emph{in general}.
\item It is impractical to run any existing MTE system on ``large'' inputs,
  containing more than a few dozen definitions.
\end{itemize}

To solve both of these problems we instead \emph{sample} a subset of
definitions. Given a sample size, we choose a subset of that many definitions,
and provide those definitions as input to our MTE tool. We generate a
corresponding ground truth by selecting those theorems from the corpus which
``depend on'' (contain references to) \emph{only} the definitions in that
sample. Transitive dependencies aren't required (e.g. a theorem which only
references \texttt{times} does not depend on \texttt{plus}, even if
\texttt{plus} occurs in the definition of \texttt{times}).

Unfortunately, uniform sampling of definitions gives rise to a lottery: for a
given sample size, increasing the size of the corpus (which, as detailed above,
is advantageous) makes it less likely that a chosen sample will contain all of
a theorem's dependencies. The majority of such samples would hence have an empty
ground truth, and thus $0$ precision and undefined recall \emph{independent} of
what the MTE tool produces! This is clearly undesirable as an evaluation
method.

Instead, we only allow a sample if it contains all dependencies of at least one
theorem. We could do this using rejection sampling, but it is more efficient to
pick a theorem, weighted in proportion to their number of dependencies (ignoring
those with too many dependencies to fit in our sample). We use those
dependencies as our sample, padded up to the required size with definitions
chosen uniformly from the remainder.

@@

Given a sample of definitions and a corresponding ground truth, the actual
execution of the MTE system proceeds as in prior work. We must translate the
chosen definitions into the particular input format of the tool under study,
then we time the execution, with a timeout after e.g. 5 minutes.

In our experiments we have found that memory usage is also an important part of
a system's performance, but rather than complicating our analysis with an extra
dimension, we instead set a limit (as large as practical) and terminate program
runs which cross it.~\footnote{We could treat the hardware's RAM capacity as our
  limit, allow the operating system to swap any overflow to disk, and wait for
  the inevitable slowdown to trigger the timeout. This requires no work to
  implement, but wastes a lot of time waiting for runs which could otherwise be
  terminated early.} This is in line with the expected usage of these systems:
either there is enough memory, or there isn't; implementations shouldn't be
penalised for making use of available resources.

This approach produces a single runtime, precision and recall value for each
sample. We propose two methods for analysing this data: \emph{summarising}
the performance of an MTE system, or \emph{comparing} the performance of
\emph{two} MTE systems.

@@

Each sample is only explored once, so that we cover as many independent samples
as possible to better estimate how a system's performance generalises to unseen
inputs. To combine these data into an aggregate summary depends on what we are
interested in measuring.

One general question we might ask is how a system's performance scales with
respect to the input size (the number of definitions in the theory). This is
straightforward to measure by varying the sample size, but we need some way to
combine the results from samples of the same size.

The precision and recall for multiple samples $S$ can be combined in two ways:
finding their mean value (the \emph{average of the ratios}) or by summing the
numerators and denominators (the \emph{ratio of the averages}). The former gives
the expected precision and recall for a run, which is of direct relevance to
users of these tools, and is hence our preferred measure. The latter assigns
more weight to theories which generate more conjectures, but we avoid this since
there is no \emph{a priori} justification for considering such theories as
``better''.

We summarise the runtimes by choosing their median, as this is more robust
against long-running outliers, and hence represents performance for a
``typical'' theory.

Another advantage to our statistical approach is that we can compute the
\emph{spread} of our data, which is unknown in the previous one-off
evaluations. In particular, we might be concerned that such previous evaluations
used ``popular'' theories, like those involving natural numbers, which may be
unrepresentative of typical usage: we should not expect a typical user's theory
(e.g. a software library) to be as mathematically rich as number theory!

@@

To encourage competition in the field, it is important to compare the relative
performance of different MTE systems on the same task. Since the aggregate
statistics in the above summaries do not include details of specific runs, any
comparison based on them would have very low statistical power, and therefore it
is appropriate to use an alternative approach when comparing.

>>>>

% Look carefully at each tool's (paper's) eval section

@@

% Lay out the field of TE: motivation, tools, evaluation, problems...

@@

We attempt to solve this discrepancy, at least for the foreseeable future, by
defining a standard, unambiguous benchmarking approach with which to compare
the conjecture generation of MTE tools. Our contributions include:

\begin{itemize}
\item A general methodology for benchmarking conjecture generation.
\item Resolving the issue of ``interestingness'' through the use of
  theorem-proving benchmarks as a ground-truth.
\item A specific instantiation of this methodology, using the Tons of Inductive
  Problems benchmark as a corpus.
\item Automated tooling to perform this benchmarking.
\item Application of our methodology to the QuickSpec and IsaCoSy MTE tools,
  and a comparison and discussion of the results.
\end{itemize}

@@

A major challenge for both ATF and MTE is choosing how to narrow down
the set of generated conjectures to those deemed ``interesting'', since this is
an imprecise term with many different interpretations. For example, all existing
approaches agree that simple tautologies are ``uninteresting'', but differ when
it comes to more complex statements.  Colton \etal{} give a survey of tools for
theory formation and exploration, and their associated notions of
``interestingness'' for concepts and conjectures~\cite{colton2000notion}. Six
notions are identified, which are applied to conjectures as follows:

@@

{\bf Applicability} depends on the number of models in which a
conjecture holds. The tool of \Bagai{} conjectures the non-existence of
objects, and hence favours statements with zero applicability. Other tools treat
applicability as a positive aspect: the more applicable the statement, the more
interesting it is.

{\bf Comprehensibility} depends on the complexity of a statement. Simpler
statements are considered more interesting, hence many of the tools explore
simpler/smaller statements before complex/larger ones, to more efficiently find
those which are interesting.

@@

\begin{table}
  \centering
  \begin{tabular}{ |l|l|c|c|c|c|c|c| }
    \hline
    \multirow{2}{*}{\textbf{Program}}                      &
    \multirow{2}{*}{\textbf{Conjecture Types}}             &
    \multicolumn{6}{c|}{\textbf{Interestingness Measures}} \\ \hhline{~~------}
    \tRow{          &                     & \iE & \iN & \iS & \iA & \iC & \iU}
    \tRow{AM        & \tIFF, \tIMP, \tNE  &   X &   X &   X &   X &   X &   X}
    \tRow{GT        & \tIFF, \tIMP, \tNE  &   X &   X &   X &   X &   X &   X}
    \tRow{Graffiti  & \tINE               &   X &   X &   X &     &   X &   X}
    \tRow{\Bagai{}  & \tNE                &     &   X &     &   X &   X &    }
    \tRow{HR        & \tIFF, \tIMP, \tNE  &   X &   X &   X &   X &   X &   X}
    \tRow{QuickSpec & \tEQ                &   X &   X &     &   X &   X &    }
    \tRow{Speculate & \tCON\ \tEQ / \tINE &   X &   X &     &   X &   X &    }
    \tRow{IsaCoSy   & \tEQ                &   X &   X &     &   X &   X &   X}
    \tRow{IsaScheme & \tEQ                &   X &   X &     &   X &   X &   X}
  \end{tabular}
  \caption{Classification of MTE tools from~\cite{colton2000notion}, extended
    to include four more recent tools. The interestingness measures are
    \iE{}mpirical plausibility, \iN{}ovelty, \iS{}urprisingness,
    \iA{}pplicability, \iC{}omprehensibility (low complexity) and \iU{}tility.}
  \label{table:colton}
\end{table}

@@

We summarise the criteria in Table 1 and state which were used in key
ATF/MTE tools. The entries for AM, GT, Graffiti, \Bagai{} and HR are based
on this survey (the latter is Colton's own tool). We extend this analysis to
some more recent tools in the remaining rows (based on our understanding of
their function): QuickSpec~\cite{QuickSpec},
Speculate~\cite{braquehais2017speculate},
IsaCoSy~\cite{Johansson.Dixon.Bundy:conjecture-generation}
and IsaScheme~\cite{MontanoRivas2011}.

The latter tools are described in more detail in section~\ref{section:mte},
but all use testing to check for counterexamples, which ensures results are
empirically plausible and applicable (conditions are satisfiable, types are
inhabited, etc.). They also use a small-to-large search order, ensuring more
easily-comprehensibile conjectures are explored first.

@@

This diversity of approaches makes it difficult to compare the existing
evaluations of these tools directly. In particular, evaluation methods created
for one tool might not make sense for another, due to assumptions made about
the algorithms' operation. For example, the novelty filters used by QuickSpec
and Speculate are applied to the whole output set, guaranteeing that no
special-cases will appear; yet we cannot assume this in the case of IsaCoSy,
since its constraint solver allows special cases if they're found \emph{before}
the general case (it does not go back and discard previous results).

With such ambiguous and varied goals, approaches and assessment criteria it is
difficult to compare MTE tools in a quantitative way, and hence to form some
measure of ``progress'' for the field. Our proposed benchmarking methodology
attempts to remedy this by setting a clear target, hopefully in line with the
myriad goals of existing researchers, whilst being generally applicable to
tools of this type.

@@

In this section we put the Mathematical Theory Exploration approach
into context both historically, and with respect to related approaches
such as in Automated Theory Formation tools.

@@

The first automated mathematical tool, the mechanical calculator
(known as the Pascaline), was an adding machine that could perform
additions and subtractions directly and multiplication and divisions
by repetitions, and was conceived by Pascal in 1642 while reorganising
tax revenues~\cite{d'ocagne}. Subsequent early tools include
M\"uller's universal calculating machine in 1784, which he built for
the purpose of calculating and printing numerical tables: he invented
this when he had to check and recalculate some tables relating to the
volumes of trees~\cite[p. 65]{lindgren}. Thirty seven years later,
Babbage invented his famous difference engine: an automatic,
mechanical calculator designed to tabulate polynomial functions. This
was inspired by a flawed table of logarithms and the idea that
machines would be quicker and more reliable~\cite{bowden}. It is
interesting to note background and motivation: while Pascal and
Babbage were mathematicians, with an interest in engineering, M\"uller
was an engineer with an interest in mathematical knowledge. All three
tools were conceived as an aid to mathematicians, as well as
scientists, accountants and surveyors.

@@

We limit our discussion to applications in mathematics, although there are
similar approaches being applied to the task of Computational \emph{Scientific}
Discovery~\cite{king2004functional,Williams20141289,schmidt2009distilling}.
Scientific knowledge relies on inductive reasoning and experimental testing,
which (in principle) are not so important to a mathematical theory. However,
these techniques \emph{do} play an important role in many mathematical theory
formation and exploration tools: in particular, inductive reasoning allows a
tool to conjecture statements which may be too difficult for it to deductively
prove; whilst experimental testing, e.g. by checking universally-quantified
statements against some specific values, is a simple way to quickly discard many
obvious falsehoods generated by powerful but imperfect algorithms (as opposed
to, say, limiting ourselves to verified theorem generators).

\subsection{Automated Theory Formation}
\label{section:atf}

@@

% 4. example tools - AM, HR, ...
When Lenat built the AM system~\cite{lenat:77}, there were systems
which could define new concepts for investigation, such as those
described in~\cite{winston}, and systems which could discover
relationships among known concepts, such as Meta-Dendral~\cite{buchanan:75}. No
system could perform both of these tasks: Lenat saw this as the next step. AM
was designed to both construct new concepts and conjecture relationships between
them; fully automating the cycle of discovery in mathematics. Lenat describes
this as follows:

@@

We believe that ATF and MTE share many goals and would benefit from a
closer alignment: both fields are highly specialised and by joining
forces (while being explicit about differences) the techniques being
developed would have greater impact. This closer alignment is already
beginning to take place. The main commonalities between MTE and ATF
are that both are interested in the same aspects of mathematical
thinking, aiming to automatically construct and evaluate elements of a
mathematical theory, including concepts, conjectures, theorems, axioms
and examples. Both strands contrast themselves with Automated Theorem
Proving. The main historical differences between these two approaches
seem to be that in MTE:

\begin{itemize}
\item the main proponents define themselves as mathematicians (they
  hold a PhD and have experience in research mathematics);
\item the primary motivation is to support mathematicians;
\item systems tend to be user-interactive;
\item systems tend to be specific to mathematics,
\end{itemize}

@@

There are three aspects to consider when evaluating a tool for conjecture
generation:

\begin{enumerate}
\item The quality of each individual conjecture, which we will refer to as their
  ``interestingness''.
\item The quality of the \emph{set} of generated conjectures, which is important
  for criteria like novelty, as well as to assess consistency.
\item Resources used by the tool, to find the balance struck between output
  quality and time taken.
\end{enumerate}

We make the assumption that each tool is intended to be \emph{general}, and
hence isn't hard-coded or fine-tuned to only perform well in a narrow domain.
Examples where this does \emph{not} hold include the natural number heuristics
of AM discussed in section~\ref{section:atf}, as well as machine learning models
trained on some specific dataset (such as this benchmark).

Evaluation of new tools and techniques for conjecture generation tends to be
\emph{ad hoc} and hence difficult to compare, whilst comparative surveys (such
as Colton's) are descriptive and qualitative rather than providing direct
numerical measurements.

One quantitative approach which has gained traction is \emph{precision/recall
  analysis}. This has been applied to IsaCoSy, IsaScheme and HipSpec, the latter
providing a direct comparison of these
tools~\cite{claessen2013automating}.\footnote{Unfortunately this comparison only
  reports prior results from the other tools, rather than reproducing them.
  This makes the numbers less comparable, since each tool was tested with
  slightly different definitions (e.g. whether or not the exponential and
  maximum functions were included).} To measure precision and recall we first
choose a set of definitions, such as those in figure \ref{figure:list_theory},
and a \emph{ground truth} to treat as the 'ideal' set of conjectures we would
like a tool to produce for those definitions. Each tool is run and their outputs
are compared to the ground truth. To score 100\% on precision and recall, all of
the conjectures which appear in the ground truth must be generated, and nothing
else:

\begin{itemize}
\item \emph{Precision} is the proportion of a tool's output which appears in
  the ground truth. This penalises overly-liberal tools which output large
  numbers of conjectures in the hope that some turn out to be ``good''.
\item \emph{Recall} is the proportion of the ground truth which appears in the
  tool's output. This penalises overly-conservative tools which generate very
  few conjectures as a way to avoid ``bad'' ones.
\end{itemize}

Precision and recall act as quality tests for sets of conjectures, fulfilling
one of our requirements. They also provide a coarse quality assessment for
individual conjectures: either a conjecture appears in the ground truth set or
it doesn't.

The analyses in~\cite{claessen2013automating} took their definitions and ground
truths from the standard library of the Isabelle theorem prover: since the
library authors have gone to the effort of stating, proving and including these
theorems in every copy of their software, this is a good indication that they
are interesting.

Two tests were performed: one on a theory of natural numbers and one on a theory
of lists (whose definitions are shown in figure~\ref{figure:list_theory}). These
tests are limited both in number (it is difficult to draw general conclusions
from two datapoints) and by the small size of these theories (e.g. the natural
number theory from~\cite{Johansson.Dixon.Bundy:conjecture-generation} contains
only 4 definitions and 12 theorems).

We believe that this is a viable approach to drawing meaningful, objective
conclusions about such tools, but larger data sets are needed to judge
performance in different, especially \emph{novel}, domains (which, after all, is
the purpose of such tools). In addition, we might be concerned that such
``popular'' theories may be unrepresentative of typical usage. For example, we
would not expect the library functions in a typical verification problem to be
as mathematically rich as number theory.

@@

Our main contribution is a benchmarking methodology, shown in figure
\ref{figure:flow_chart}, which both generates a large definition/ground-truth
corpus, and provides a scalable, statistical approach to evaluating MTE tools
using this corpus. We follow a precision/recall approach similar to prior work,
with the main difference being the source of definitions and ground truths: we
take existing problem sets designed for automated theorem proving, and adapt
their content for use in the conjecture generation setting.

@@

Automated theorem proving is an active area of research, with large problem sets
and regular competitions to prove as much as possible, as fast as
possible~\cite{pelletier2002development}. These problem sets are an opportunity
for MTE, as their definitions and theorems can be used as a corpus in the same
way that Isabelle libraries have been used in the past.

@@

\begin{itemize}
\item For each problem, there should be a clear distinction between the theorem
  to be proved and the definitions involved, such that the two can be easily and
  meaningfully separated. This rules out problem sets like those of
  SMT-COMP~\cite{barrett2005smt}, where many problems involve uninterpreted
  functions, whose behaviour is \emph{implicit} in the logical structure of the
  theorem statement but not separately \emph{defined}.
\item Definitions should be translatable into a form suitable for the MTE tools
  under study. Our application translates to Haskell and Isabelle, which also
  benefits from having definitions be strongly typed.
\item The problem set should be relevant to the desired domain. Our focus on
  functional programming requires higher-order functions and inductively defined
  datatypes, which rules out first-order languages/logics (such as
  TPTP~\cite{sutcliffe2009tptp}).
\item Since it will act as our ground truth, the problem set should ideally
  contain every ``interesting'' conjecture involving its included
  definitions. Realistically, we should aim for each definition to appear in
  \emph{many} theorem statements; rather than each problem having unique
  definitions.
\item The problem set should be as large as possible, for robustness of the
  resulting statistics.
\end{itemize}

Once such a problem set has been chosen, we must separate the definitions
referenced by the theorems from the theorem statements themselves. The former
will be used as input to the tools, whilst the latter form a ground truth corpus
against which to compare the output.

It is important to ensure that there are no duplicate definitions: we are only
concerned with the \emph{logical} content of the input, not the more arbitrary
aspects like the names of functions. For example, consider a problem set which
includes a commutativity theorem for a \texttt{plus} function, and an
associativity theorem for an \texttt{add} function, where the definitions of
\texttt{plus} and \texttt{add} are $\alpha$-equivalent. We would expect an MTE
tool to either conjecture commutativity and associativity for \emph{both}
functions, or for \emph{neither} function, since they're logically equivalent.
Yet a na\"ive precision/recall analysis would treat commutativity of
\texttt{add} and associativity of \texttt{plus} as \emph{uninteresting}, since
they don't appear in the ground truth.

@@

We could, in theory, send these de-duplicated definitions straight into an MTE
tool and use the updated theorem statements as the ground truth for analysis.
However, this would cause two problems:

\begin{itemize}
\item The result would be a single data point, which makes it difficult to
  infer performance \emph{in general}.
\item It is impractical to run existing MTE tools on ``large'' inputs,
  containing more than a few dozen definitions.
\end{itemize}

To solve both of these problems we instead \emph{sample} a subset of
definitions.\footnote{This could be done randomly, but for reproducibility we
  use a deterministic order (based on cryptographic hashes, to discourage
  ``cheating'').} Given a sample size, we choose a subset of that many definitions,
and provide only those as input to the tool. We generate a corresponding
ground truth by selecting those theorems from the corpus which ``depend on''
(contain references to) \emph{only} the definitions in that sample. Transitive
dependencies aren't required (e.g. a theorem which only references
\texttt{times} does not depend on \texttt{plus}, even if \texttt{plus} occurs in
the definition of \texttt{times}).

Unfortunately, uniform sampling of definitions gives rise to a lottery: for a
given sample size, increasing the size of the corpus (which provides better
statistics) makes it less likely that a chosen sample will contain all of
a theorem's dependencies. The majority of such samples would hence have an empty
ground truth, and thus $0$ precision and undefined recall \emph{independent} of
the tool's output! This is clearly undesirable as an evaluation method.

Instead, we only allow a sample if it contains all dependencies of at least one
theorem. We could do this using rejection sampling, but it is more efficient to
pick a theorem, weighted in proportion to their number of dependencies and
ignoring those with more dependencies than our sample size. That theorem's
dependencies become our sample, padded up to the required size with uniform
choices from the remaining definitions.

@@

Given a sample of definitions and a corresponding ground truth, the actual
execution of the MTE tool proceeds as in prior work. We must translate the
chosen definitions into the required input format, then we time the execution
with a timeout (e.g. 5 minutes).

In our experiments we have found that memory usage is also an important part of
a tool's performance, but rather than complicating our analysis with an extra
dimension, we instead allow programs to use as much memory as they like, and
either get killed by the operating system or slow down so much from swapping
that they time out. This is in line with the expected usage of these tools:
either there is enough memory, or there isn't; implementations shouldn't be
penalised for making use of available resources.

To calculate precision and recall, the conjectures generated by each tool, as
well as the ground-truth theorems, need to be parsed into a common format and
compared syntactically after normalising away ``irrelevant'' details. We
consider variable naming to be irrelevant, which can be normalised by numbering
free variables from left to right and using de Bruijn indices for bound
variables. We also consider the left/right order of equations to be irrelevant,
which we normalise by choosing whichever order results in the
lexicographically-smallest expression.

Other than $\alpha$-equivalence and equation symmetry, we specifically
\emph{ignore} logical relationships between syntactically distinct statements,
such as one equation being implied by another. Whilst logically sound,
second-guessing the ground truth in this way would harm other aspects which
influence interestingness (for example, a more general statement is more widely
applicable, but it might also be harder to comprehend).

This procedure gives us a single runtime, precision and recall value for each
sample. We propose two methods for analysing this data: \emph{summarising}
the performance of a single MTE tool and \emph{comparing} the performance of two
tools.

@@

Each sample is only explored once by each tool, so that we cover as many
independent samples as possible to better estimate how a tool's performance
generalises to unseen inputs. How we combine these data into an aggregate
summary depends on what we are interested in measuring.

One general question we might ask is how a tool's performance scales with
respect to the input size (the number of given definitions). This is
straightforward to measure by varying the sample size, but we need some way to
combine the results from samples of the same size.

We can summarise runtimes by choosing their median, as this is more robust
against long-running outliers, and hence represents performance for a
``typical'' input of that size. Unlike previous one-off evaluations, we can also
compute the \emph{spread} of our data, for example with the inter-quartile
range.

The precision and recall for multiple samples can be combined in two ways. We
can calculate their mean value, known as the \emph{average of the ratios}, to
find the expected precision and recall for runs of this size. Alternatively, we
can sum their numerators and denominators to get the \emph{ratio of the
  averages}: this normalises the size of the generated and ground truth sets,
letting us model precision and recall as simple Bernoulli processes.

@@

To measure progress and encourage competition in the field, it is important to
compare the relative performance of different tools on the same task. Since the
aggregate statistics in the above summaries have obscured the details of
specific runs, any comparison based on them (such as comparing average recall)
would have very low statistical power. More direct comparisons can be made using
\emph{paired} tests, since for each individual sample we have measurements for
\emph{both} tools.
